#Reproducible Research: Peer Assessment 1
##Author: Nick Grab

###Loading and Preprocessing the Data

Since this assingment has specific questions around missing data points, there was not much preprocessing required, beyond downloading the data.

```{r}
temp <- tempfile()
fileURL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, temp)
activityData <- read.csv(unz(temp,"activity.csv"))
```
###What is mean total number of steps taken per day?

To answer this question I used the ddply function in the plyr package, summarizing total, average, and median steps by calendar date.  Missing values are ignored and values are reported to the nearest full step.
```{r warning=FALSE}
library(plyr)
stepsData <- ddply(activityData, "date", summarize,
                   total = sum(steps, na.rm = TRUE),
                   average = round(mean(steps, na.rm = TRUE)),
                   median = round(median(steps, na.rm = TRUE)))
```
I then created a histrogram of total daily steps:
```{r}
hist(stepsData$total)  
```

**Summary table of average and median steps by calendar date.**  Rows with non-numeric values (i.e. NaN, NA) are dates where activity data are wholly missing; this issue addressed further in the *Inputting missing values* section below.
```{r}
stepsData[,c(1,3,4)]
```
###What is the average daily activity pattern?

To answer this question I used the ddply function in the plyr package, summarizing average daily steps by interval.  Missing values are ignored and values are reported to the nearest full step.
```{r}
intervalData <- ddply(activityData, "interval", summarize,
                   average = round(mean(steps, na.rm = TRUE)),
                   median = round(median(steps, na.rm = TRUE)))
```
The data was then plotted as time series:
```{r}
plot(intervalData$interval,intervalData$average, type = "l")
```

**The maximum average number of steps taken was `r max(intervalData$average, na.rm = TRUE)`, and occurs in interval `r intervalData[(intervalData$average == max(intervalData$average, na.rm = TRUE)),1]`.**

###Inputing missing values
There are `r a <- count(complete.cases(activityData)); a$freq[1]` rows with missing data in this data set.  This number was found by counting the number of complete cases (i.e. records without NA).
```{r}
a <- count(complete.cases(activityData))
a$freq[1]
```
I decided to replace missing data with the median step of the corresponding interval.  The median value was chosen over the average to avoid skewing data by outliers.
```{r}
filledData <- activityData
for (i in 1:nrow(activityData)){
  if(is.na(activityData$steps[i])){
    filledData[i,1] <- intervalData[(intervalData$interval == activityData$interval[i]),3]
  } else{
    filledData[i,1] <- activityData$steps[i]
  }
}
```

Again using the ddply function found in the plyr package, I summarized total, average, and median steps by calendar date.  Values are reported to the nearest full step.
```{r}
filledSteps <- ddply(filledData, "date", summarize,
                   total = sum(steps, na.rm = TRUE),
                   average = round(mean(steps, na.rm = TRUE)),
                   median = round(median(steps, na.rm = TRUE)))
```
I then created a histrogram of total daily steps:
```{r}
hist(filledSteps$total)
```

**Summary table of average and median steps by calendar date.**  Beside the fact that the table no longer contains non-numeric entries, by replacing the missing data with the median we've changed the sample size and therefore changed the averages.
```{r}
filledSteps[,c(1,3,4)]
```

###Are there differences in activity patterns between weekdays and weekends?

Using the weekdays function, I first identified each rows weekday name.  I then loopd through the dataframe another time and tagged all rows with "Saturday" or "Sunday" weekday names as "weekend"; all others were tagged as "weekday"
```{r}
filledData$day <- weekdays(as.Date(filledData$date,'%Y-%m-%d'))
for (i in 1:nrow(filledData)){
  if (filledData$day[i] == "Saturday" | filledData$day[i] == "Sunday"){
    filledData$weekPart[i] <- "weekend"
  } else {
    filledData$weekPart[i] <- "weekday"
  }
}
```
Again using the ddply function found in the plyr package, I summarized average steps by week part (i.e. weekend or weekday) and interval.  Values are reported to the nearest full step.
```{r}
intervalWeek <- ddply(filledData, c("weekPart","interval"), summarize,
                   average = round(mean(steps, na.rm = TRUE)))
```

Data was then plotted as time series in a panel plot:
```{r}
library(lattice)
xyplot(intervalWeek$average ~ intervalWeek$interval | intervalWeek$weekPart, 
       ylab = "Average No. Steps", 
       xlab = "Time [5 min. interval]",
       type = "l")
```
